Minimal Positive Influence of Helpline Efforts on Reduction of Homeless Death in Toronto

### 2024-03-16 paper/paper.qmd
finished final touches, waiting for posit cloud to become available for rendering

Final dataset only includes data from November 2020 to June 2023
(@tbl-death-count). The rationalization behind the action will be
explained in @sec-data-diff-datasets.

@@

$y_i|\mu_i$, $\sigma \sim \mbox{Normal}(\mu_i, \sigma)$

$\mu_i = \alpha + \phi_i$

$\alpha \sim \mbox{Normal}(0, 2.5)$

$\phi \sim \mbox{Normal}(0, 2.5)$

$\sigma \sim \mbox{Exponential}(1)$

@@

$y_i|\mu_i$, $\sigma \sim \mbox{Normal}(\mu_i, \sigma)$

$\mu_i = \alpha + \beta_i + \gamma_i$

$\alpha \sim \mbox{Normal}(0, 2.5)$

$\beta \sim \mbox{Normal}(0, 2.5)$

$\gamma \sim \mbox{Normal}(0, 2.5)$

$\sigma \sim \mbox{Exponential}(1)$

@@

-   Under the Gaussian regression model, @fig-total-coded illustrates a
    fitted line that estimates the linear relationship between the
    number of total calls coded and homeless death count in Toronto. The
    graph suggest that there exist a net negative correlation between
    the two variable. According to the graph, homeless death count
    decrease as total number of calls coded increase. The significance
    of this relationship is discussed below in in the model summary
    table.

@@

-   The model summary table, @tbl-modelresults, displays the coefficient
    values of the predictor variables against response variable:
    homeless death count in Toronto. Coefficients of predictors indicate
    their influence against the response variable. For example, the
    column on the left shows that, under Model 1, every 1 additional
    helpline call coded and decrease the homeless death count by 0.0003.
    The ***+p \<...*** section at the bottom of the table represents
    statistically significance of the coefficient values. As indicated
    by the ***+p\<...*** legend, the coefficients with 1 \*, coefficient
    is significant at 95% confidence level; for \*\*, coefficient is
    significant at 99% confidence level, and so on. However, as we can
    see from the table below, no \* exists for the coefficients in both
    models.

@@

-   On average, for every increase in calls coded by the helpline, there
    is a 0.0003 unit decrease in homeless death in Toronto. There is an
    overall reduction, albeit very small, in homeless death as the
    number of total coded calls increase. This implies that the helpline
    need to take at minimum 3333 calls per month to save 1 homeless life
    in Toronto. On average, the helpline takes 14828.59 calls per month,
    meaning that 4.45 lives are saved on average.
-   4.45 is not a significant number in both a practical and statistical
    sense. To improve the return on total number of calls coded, we
    attempted to identify promising call codes that contribute to the
    net positive reduction in homeless death. Out of the 13 different
    call codes, *Referral to Shelter* and *Information on Homeless
    Prevention* are best suited as individual efforts of measurement for
    positive impacts because they provide a count of the number of
    callers provided with directions or advice.
-   On average, for every homeless caller referred to shelter, there is
    a 0.0059 unit **increase** in homeless death in Toronto. Though not
    significant on a statistical level, this trend is still very
    counter-intuitive discouraging. In the case where the correlation is
    more significant, we would want to further analyze factor related to
    referrals to shelter. Accuracy on the current status of shelters,
    level of detail on instructed, and misjudgment of shelter
    eligibility are possible factors of examination. However, since the
    effect of this call code is statistically insignificant, it is also
    likely to be just a fluke of observation.
-   For every caller provided with information on homeless prevention
    information, there is an average decrease of 0.0048 death count in
    Toronto.

## Implication of Insignificance: More Qeustions than Answers

-   Homeless life is not easy, data shows that a total of 522 homeless
    individual died in Toronto between November 2020 to June 2023.

-   The Central Intake Line is created with the intend to help homeless
    individuals in need. It provides a 24/7 platform for individuals to
    receive government help.

-   The insignificance of its influence suggests that the helpline has
    not fulfilled its intended purpose, though it is also important to
    note that having insignificant effect on reducing death count does
    not mean the helpline has not contributed to the well-being of
    Toronto's homeless population in other ways.

-   This makes us question

@@

Weaknesses: very small sample size, only 32 observations Next steps:
investigate into the possible factors that caused a net zero impact on
reduction of homeless death count and propose possible solution using
data.

@@

# Additional data details

@@

In @fig-total-coded-model-ppcheckandposteriorvsprior-1 we implement a
posterior predictive check. This shows...

In @fig-total-coded-model-ppcheckandposteriorvsprior-2 we compare the
posterior with the prior. This shows...

@@

In @fig-referred-informed-model-ppcheckandposteriorvsprior-1 we
implement a posterior predictive check. This shows...

In @fig-referred-informed-model-ppcheckandposteriorvsprior-2 we compare
the posterior with the prior. This shows...

@@

@fig-total-coded-model-stanareyouokay-1 is a trace plot. It shows...
This suggests...

@fig-total-coded-model-stanareyouokay-2 is a Rhat plot. It shows... This
suggests...

@@

@fig-referred-informed-model-stanareyouokay-1 is a trace plot. It
shows... This suggests...

@fig-referred-informed-model-stanareyouokay-2 is a Rhat plot. It
shows... This suggests...

>>>>

Final dataset only includes data from November 2020 to June 2023
(@tbl-death-count). The rationalization behind the action will be
explained in @sec-combined-data.

@@

$$y_i|\mu_i, \sigma \sim \mbox{Normal}(\mu_i, \sigma)$$

$$\mu_i = \alpha + \phi_i$$

$$\alpha \sim \mbox{Normal}(0, 2.5)$$

$$\phi \sim \mbox{Normal}(0, 2.5)$$

$$\sigma \sim \mbox{Exponential}(1)$$

@@

$$y_i|\mu_i, \sigma \sim \mbox{Normal}(\mu_i, \sigma)$$

$$\mu_i = \alpha + \beta_i + \gamma_i$$

$$\alpha \sim \mbox{Normal}(0, 2.5)$$

$$\beta \sim \mbox{Normal}(0, 2.5)$$

$$\gamma \sim \mbox{Normal}(0, 2.5)$$

$$\sigma \sim \mbox{Exponential}(1)$$

@@

Under the Gaussian regression model, @fig-total-coded illustrates a
fitted line that estimates the linear relationship between the number of
total calls coded and homeless death count in Toronto. The graph suggest
that there exist a net negative correlation between the two variable.
According to the graph, homeless death count decrease as total number of
calls coded increase. The significance of this relationship is discussed
below in in the model summary table.

@@

The model summary table, @tbl-modelresults, displays the coefficient
values of the predictor variables against response variable: homeless
death count in Toronto. Coefficients of predictors indicate their
influence against the response variable. For example, the column on the
left shows that, under Model 1, every 1 additional helpline call coded
and decrease the homeless death count by 0.0003. The ***+p \<...***
section at the bottom of the table represents statistically significance
of the coefficient values. As indicated by the ***+p\<...*** legend, the
coefficients with 1 \*, coefficient is significant at 95% confidence
level; for \*\*, coefficient is significant at 99% confidence level, and
so on. However, as we can see from the table below, no \* exists for the
coefficients in both models.

@@

-   On average, for every increase in calls coded by the helpline, there
    is a 0.0003 unit decrease in homeless death in Toronto. There is an
    overall reduction, albeit very small, in homeless death as the
    number of total coded calls increase. This implies that the helpline
    need to take at minimum 3333 calls per month to save 1 homeless life
    in Toronto. On average, the helpline takes 14828.59 calls per month,
    meaning that 4.45 lives are saved on average.
-   4.45 is not a significant number in both a practical and statistical
    sense. To improve the return on total number of calls coded, we
    attempted to identify promising call codes that contribute to the
    net positive reduction in homeless death. Out of the 13 different
    call codes, *Referral to Shelter* and *Information on Homeless
    Prevention* are best suited as individual efforts of measurement for
    positive impacts because they provide a count of the number of
    callers provided with directions or advice. For every caller
    provided with information on homeless prevention information, there
    is an average decrease of 0.0048 death count in Toronto. On the
    other hand, for every homeless caller referred to shelter, there is
    a 0.0059 unit of average **increase** in homeless death in Toronto.
    This trend is very counter-intuitive and discouraging at first
    glance. In the case where the correlation is more significant, we
    would want to further analyze factor related to referrals to
    shelter. Accuracy on the current status of shelters, level of detail
    on instructed, and misjudgment of shelter eligibility are possible
    factors of examination. However, since the effect of this call code
    is statistically insignificant, it is also likely to be just a fluke
    of observation.

## Factors Contributing to Insignificance

Homeless life is not easy, data from @Homeless_Death_Count shows that a
total of 522 homeless individual died in Toronto between November 2020
to June 2023. The Central Intake Line is created with the intend to help
homeless individuals in need. It provides a 24/7 platform for
individuals to receive government help. The insignificance of its
influence suggests that the helpline has not fulfilled its intended
purpose, though it is also important to note that having insignificant
effect on reducing death count does not mean the helpline has not
contributed to the well-being of Toronto's homeless population in other
ways. The following two subsections discusses two of the possible
factors that contributes to the insignificance of the helpline's
influence on the reduction of homeless death in Toronto: inaccessibility
to phone lines and substitute calling options.

### Inaccessibility of Phone Lines

Difficulty in accessing the helpline may contribute to the
insignificance of the influence the helpline has on the reduction of
homeless death in Toronto. In order to access the resources provided by
the helpline, a homeless individual is required to have access to one of
the following three tools: mobile phone with valid phone plan, free
city-run phone, or payphone. Financially, the most expensive option is
the mobile phone. In 2024, the cheapest calling plan in Toronto starts
at \$15 before tax, discouraging the less wealthy proportion of the
homeless population. The cheapest alternative is the city-run free phone
lines accessible in several City buildings and Universities across
Toronto. According to @localwiki, there are currently 9 locations in
downtown Toronto offering access to free phone lines. However, the free
phone line access is only available during the operation hours of the
respective buildings, meaning that free calls are not available 24/7.
This leaves the last option, the affordable and readily available local
payphones. Currently, Toronto payphones cost 50 cents per call, which is
definitely more financially reasonable for the homeless population.
However, numerous articles through the years have pointed out the steady
decline of functional payphones in Toronto. The popularization of
smartphones in the last decade contributes to the decrease in the demand
of payphones, subsequently causing more and more abandoned or
un-repaired payphone stations. This difficulty in accessing the helpline
will decrease the number of callers, subsequently lowering the influence
of the helpline on reducing homeless death in Toronto.

### Substitute Calling Options

The Central Intake Line is one of the biggest city-run homeless
helplines in Toronto; however, it is not the only one. Using keywords
“homeless helpline Toronto”, search results such as City of Toronto Line
(311) and Assaulted Women’s Helpline also appear in the Google search
engine. The existence of multiple helplines will cause specific problems
to go to specific channels, reducing the total percent of homeless
population that choose to use the Central Intake Line inquired in this
paper. Similar to the previous factor, reduction in the total number of
calls per month reduces the average influence of the helpline on the
reduction of homeless death. Other potential factors that may contribute
to the insignificance of influence are: changes in city policies, lack
of promotion through non-internet platforms, etc.

@@

The biggest weakness for this paper is the small sample size of the
data, there are only 32 observations in the final data used for
analysis. This suggests that resulting outputs may not accurately
represent the situation in the real world because trends might only
exist in our observed sample. Some potential next steps are:
investigating into the proposed factors that contributes to the
insignificance of helpline influence on homeless death using real-world
data and models, as well as researching further into other potential
factors.

@@

@@

In @fig-total-coded-model-ppcheckandposteriorvsprior-1 we implement a
posterior predictive check. This shows a relatively close match between
the regression outcome from our data and the simulation created by the
posterior distribution. This suggests that our model is a relatively
good fit for our data.

In @fig-total-coded-model-ppcheckandposteriorvsprior-2 we compare the
posterior with the prior. We can see that the estimation parameters
shifts minimally after taking data into account. This suggests that good
prior parameters were set.

@@

In @fig-referred-informed-model-ppcheckandposteriorvsprior-1 we
implement a posterior predictive check. This shows a relatively close
match between the regression outcome from our data and the simulation
created by the posterior distribution. This suggests that our model is a
relatively good fit for our data.

In @fig-referred-informed-model-ppcheckandposteriorvsprior-2 we compare
the posterior with the prior. We can see that the estimation parameters
shifts minimally after taking data into account. This suggests that good
prior parameters were set.

@@

@fig-total-coded-model-stanareyouokay-1 is a trace plot. Trace plots are
often used to diagnose abnormalities of models. In
@fig-total-coded-model-stanareyouokay-1, lines are oscillating
vertically and maintains an overall horizontal trend. This suggests that
there is nothing out of the ordinary for our model.

@fig-total-coded-model-stanareyouokay-2 is a Rhat plot. Rhat plot is a
valuable tool for assessing convergence in Bayesian regression models,
helping ensure the reliability and validity of the model's inference
results. In @fig-total-coded-model-stanareyouokay-2, all data points are
close to 1. This suggest that there is nothing out of the ordinary for
our model.

@@

@fig-referred-informed-model-stanareyouokay-1 is a trace plot. Trace
plots are often used to diagnose abnormalities of models. In
@fig-referred-informed-model-stanareyouokay-1, lines are oscillating
vertically and maintains an overall horizontal trend. This suggests that
there is nothing out of the ordinary for our model.

@fig-referred-informed-model-stanareyouokay-2 is a Rhat plot. Rhat plot
is a valuable tool for assessing convergence in Bayesian regression
models, helping ensure the reliability and validity of the model's
inference results. In @fig-referred-informed-model-stanareyouokay-2, all
data points are close to 1. This suggest that there is nothing out of
the ordinary for our model.
